---
title: "Discrete Random Variables"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

```{r}
library(ggplot2)
library(dplyr)
```

---

**Scenario**: A salesman has scheduled two appointments to sell software, one in the morning and another one in the afternoon. There are two software editions available: the base edition costing Rs. 500 and the premium edition costing Rs. 1000. His morning appointments typically lead to a sale with a 30% chance while the afternoon ones typically lead to a sale with a 60% chance independent of what happened in the morning. If the morning appointment ends up in sale, the salesman has a 70% chance of selling the premium edition and if the afternoon appointment ends up in a sale, he is equally likely to sell either of the editions. Let X be the random variable representing the total Rupee value of sales. What are the different values that X can take? Calculate the probability that X takes the value 5000? 

---


```{r}
s_appointment = c(0,1)
```

 

```{r}
p_morning = 0.30
p_afternoon = 0.60
p_appointment = matrix(nrow = 2, ncol = 2, c(1-p_morning, p_morning, 1 - p_afternoon, p_afternoon), byrow = TRUE)
```

 

```{r}
s_software = c(5000,10000)
```

```{r}
p_software = matrix(nrow = 2, ncol = 2, c(0.3,0.7,0.5,0.5), byrow = TRUE)
```

 

```{r}
salesResult = function(){
  result = numeric(2)

  for (j in c(1:2)){
    result[j] = sample(s_appointment,size = 1, replace = TRUE, prob = p_appointment[j,])}
  earnings = ifelse(result[1] == 1, sample(s_software,size = 1, replace =  TRUE, prob = p_software[1,]),0) +
    ifelse(result[2] == 1, sample(s_software, size = 1, replace = TRUE, prob = p_software[2,]), 0)
  return(earnings)
}
```

```{r}
nsimulations = 1e5

 

simulatedData = replicate(nsimulations, salesResult())
```

```{r}
# prob that salesman earns 5k
mean(simulatedData == 5000)
```

```{r}
a = table(simulatedData)/nsimulations
```


```{r}
a
```

---

**Discrete random variable**: Let $X$  represent thetotal earnings from one day whic is random; that is, $X$ is a discrete random variable which can take the values $0, 500, 1000, 1500, 2000.$ The associated probabilities can be calculated as:

$\begin{align*}P(X = 0)  &= 0.28,\\ P(X = 500) &= 0.31,\\ P(X = 1000) &= 0.275,\\ P(X = 1500) &= 0.095,\\ P(X = 2000) &= 0.04.\end{align*}$

---

**Probability Mass Function (PMF)** of the random variable $X$ is denoted as $P_X(x),$ where $x$ represents the possible values that the random variable $X$ can take:

$\begin{align*}{\color{red}{P_X}}(0) &= P(X = 0)  = 0.28,\\ {\color{red}{P_X}}(500) &=P(X = 500) = 0.31,\\ {\color{red}{P_X}}(1000) &=P(X = 1000) = 0.275,\\ {\color{red}{P_X}}(1500) &=P(X = 1500) = 0.095,\\ {\color{red}{P_X}}(2000) &=P(X = 2000) = 0.04.\end{align*}$

---

```{r}
dfSales = as.data.frame(simulatedData)
colnames(dfSales) = 'Earning'
p=ggplot(data = dfSales)+geom_bar(aes(x = Earning),fill = 'blue')
p
```


```{r}
# Convert simulatedData into a dataframe
dfSales = as.data.frame(table(simulatedData))
# Add names to the columns
colnames(dfSales) = c('Earning', 'Frequency')

# Add a new column with proportion (probability) values
dfSales['Probability'] = dfSales['Frequency']/nsimulations

head(dfSales, 5)

p = ggplot(data = dfSales) + geom_col(aes(x = Earning, y = Frequency), fill='blue')
p
```

What is the salesman's expected earnings? What does it even mean to say "expected earning?" It is a single number denoted as $E[X]$ and referred to as the *expected value of* $X$ which can be calculated as follows:


---



*   *From a simulation perspective*, it is simply an average of all the simulated earnings:
$\begin{align*}E[X] &\approx 0 \times\underbrace{\frac{\text{No. of times 0 appeared}}{\text{nsimulations}}}_{\text{approximation to }P_X(0)}+500 \times\underbrace{\frac{\text{No. of times 500 appeared}}{\text{nsimulations}}}_{\text{approximation to }P_X(500)}+1000 \times\underbrace{\frac{\text{No. of times 1000 appeared}}{\text{nsimulations}}}_{\text{approximation to }P_X(1000)}+1500 \times\underbrace{\frac{\text{No. of times 1500 appeared}}{\text{nsimulations}}}_{\text{approximation to }P_X(1500)}+2000 \times\underbrace{\frac{\text{No. of times 2000 appeared}}{\text{nsimulations}}}_{\text{approximation to }P_X(2000)}.\end{align*}$


---


*   *From a theoretical perspective*, s it is defined as the weighted sum of the possible values $X$ can take with the corresponding probabilities:

$\begin{align*} E[X] &= \sum_xxP_X(x),\ \text{where }x = 0, 500, 1000, 1500, 2000\\\Rightarrow E[X] &= 0\times P_X(0)+500\times P_X(500)+1000\times P_X(1000)+1500\times P_X(1500)+2000\times P_X(2000).\end{align*}$



---

So, we see that the expected value of $X$ is approximately the long-term average of the simulated (or realized) values of $X.$

```{r}
# Expected earnings using simulated values
mean(simulatedData) # long term average of the realizations of X
    
# Expected earnings using the theoretical definition
x = c(0, 500, 1000, 1500, 2000)
p = c(0.28, 0.27, 0.315, 0.09, 0.045)
sum(x * p)
```

How much could the salesman's earnings vary from the expected value of the earning calculated in the previous cell?


---


To answer this question, let's first look at the quantity $X-E[X] = X-675.$ Note that:


*   this is also a random variable with the possible values $0-675, 500-675, 1000-675, 1500-675, 2000-675$;
*   it can be thought of as the random variable that captures the deviation of $X$ from its expected value;
*  the associated probabilities are still the same as:
\begin{array}{c|c|c}
\hline
\color{blue}X&\color{blue}{X}-\color{cyan}{E[X]} = \color{blue}{X}-\color{cyan}{675}&\color{magenta}{\text{Probability}}\\
\hline
0 & -675 & 0.28\\
500 & -175 & 0.27\\
1000 & 325 & 0.315\\
1500 & 825 & 0.09\\
2000 & 1325 & 0.045
\end{array}
---
Now, consider the quanityt $(X-E[X])^2 = (X-675)^2.$ Note that:


*   this is also a random variable with the possible values $(0-675)^2, (500-675)^2, (1000-675)^2, (1500-675)^2, (2000-675)^2$;
*   it can be thought of as the random variable that captures the squared deviation of $X$ from its expected value;
*  the associated probabilities are still the same as:
\begin{array}{c|c|c}
\hline
\color{blue}X&(\color{blue}{X}-\color{cyan}{E[X]})^2 = (\color{blue}{X}-\color{cyan}{675})^2&\color{magenta}{\text{Probability}}\\
\hline
0 & (-675)^2 & 0.28\\
500 & (-175)^2 & 0.27\\
1000 & (325)^2 & 0.315\\
1500 & (825)^2& 0.09\\
2000 & (1325)^2 & 0.045
\end{array}

---

$\color{green}{Variance}$ of the random variable $X,$ denoted as $\text{var}[X],$ is the expected squared deviation of $X$ from its expected value. In simple terms, variance of $X$ is the expected value of the random variable $(X-E[X])^2.$

$\begin{align*}\text{Var}[X] &= E[\underbrace{(X-E[X])^2}_{\text{squared deviation random variable}}]\\&=
\sum_x (x-E[X])^2\times P_X(x)\\& = (0-675)^2\times P_X(0)+(500-675)^2\times P_X(500)+(1000-675)^2\times P_X(1000)+(1500-675)^2\times P_X(1500)+(2000-675)^2\times P_X(2000)\\&=(0-675)^2\times 0.28+(500-675)^2\times 0.27+(1000-675)^2\times 0.315+(1500-675)^2\times 0.09+(2000-675)^2\times 0.045.\end{align*}$

---

$\color{green}{Standard\ deviation}$ of the random variable $X,$ denoted as $\text{SD}[X],$ is defined as the square root of its variance:

$\begin{align*}\text{SD}[X] &= \sqrt{\text{Var}[X]} = \sqrt{E\left[(X-E[X])^2\right]}\\&=\sqrt{(0-675)^2\times 0.28+(500-675)^2\times 0.27+(1000-675)^2\times 0.315+(1500-675)^2\times 0.09+(2000-675)^2\times 0.045}.\end{align*}$

```{r}
# Variance of the earnings using simulated values
# long term average of this random variable (X-E[X])^2
#simulatedData - mean(simulatedData) # simulated deviations
#(simulatedData - mean(simulatedData))^2 # simulated squared deviations
mean(simulatedData)
mean((simulatedData - mean(simulatedData))^2) # variance calculated using simulation

# Standard deviation of the earnings using simulated values
sqrt(mean((simulatedData - mean(simulatedData))^2))

print(' ')

# Expected value using the theoretical definition
sum(x*p)

# Variance of the earnings using the thoretical definition
x = c(0, 500, 1000, 1500, 2000)
p = c(0.28, 0.27, 0.315, 0.09, 0.045)
sum((x - sum(x*p))^2 * p)

# Standard deviation of the earnings using the theoretical definition
sqrt(sum((x - sum(x*p))^2 * p))
```

```{r}
x = c(0, 500, 1000, 1500, 2000)
p = c(0.28, 0.27, 0.315, 0.09, 0.045)

# Deviations from the expected value
x - sum(x*p)

# Squared deviations from the expected value
(x - sum(x*p))^2

# Expected value of the squared deviations from the expected value
sum((x - sum(x*p))^2 * p)
```




```{r}
nsimulation = 100
simulatedData = rnorm(nsimulation,mean = 74, sd = 6)
simulatedData
```


[03:14 pm] MUPPIREDDY MITHILESH REDDY               - 231057009 - MSISMPL

```{r}

nsimulations = 100

simulatedData = rnorm(nsimulations, mean = 74, sd =6)

print(simulatedData)

avg = mean(simulatedData)

mean((simulatedData-mean(simulatedData))^2)

sd = sqrt(mean((simulatedData-mean(simulatedData))^2))

```

 

```{r}

df = data.frame(simulatedData)

df

ggplot(data = df, aes(x=1:100, y=simulatedData)) + geom_point(color='yellow') + geom_line(aes(y = avg), color = 'blue') + geom_line(aes(y = avg+sd), color = 'black', linetype = 'dashed') + geom_line(aes(y = avg-sd), color = 'orange', linetype = 'dashed') + labs(x = 'Patient #', y = 'Heart Rate(BPM)', title = 'Patient HeartRate')

```

```{r}
mean(simulatedData)
```

```{r}
var(simulatedData)
```


```{r}
mean(simulatedData) - mean(simulatedData)^2
```


```{r}
sd(simulatedData)
```


```{r}

sqrt(mean((simulatedData - mean(simulatedData))^2))
```

```{r}
mean(abs(simulatedData - mean(simulatedData)) <= sd(simulatedData))
```


